---
title: "Contributing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contributing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

If you are reading this vignette you are most probably to contribute to the `{mapme.biodiversity}`
package. This is great news and we are very happy to receive Pull-Requests
extending the package's functionality! Below you will receive important in-depth
information about how to add resources and indicators to make the process
as seamless as possible for both you and the package's maintainers. Please make
sure to read and understand this guide before opening a PR. If in doubt, especially
if you feel that the framework does not support your specific use case, always
feel free to raise an issue and we will happily discuss how we can support your
ideas. If you have not already done so, make sure to read [Terminology](terminology.html)
vignette to get familiar with the most important concepts of this package.

Note that we use the [tidyverse style guide](https://style.tidyverse.org/) 
for the package. That specifically means that function and variable names should 
follow the snake case pattern. We also use the arrow assignment operator (`<-`). 
When submitting a PR that does not consistently follow the tidyverse style guide, 
the maintainers of the package might change the code to adhere to this code style
without further notice before accepting the PR. 

# Getting started

Ideally, you clone the GitHub repository via the git command in a command line
on Linux and MacOS systems or via the GitHub Desktop application on Windows.
On Linux, the command would look like this:

```bash
git clone https://github.com/mapme-initiative/mapme.biodiversity
```

We do not accept pushes to main, thus the first step would be to create
a specific branch for your extension. In this tutorial, we will pretend to 
reimplement the `soilgrids` resources and the associated `soilproperties` 
indicator, so that we create a branch reflecting this. Don't forget to check out
to the newly created branch!

```bash
git branch add-soilgrid-indicators
git checkout add-soilgrid-indicators
```
Below, we will assume that you develop your extension to the package in R Studio.
The general guidelines to follow also apply if you choose different tooling 
for your development process, however, it will not be covered in this vignette.
We assume that all R development dependencies for the state of the package when 
you used the `git clone` command are installed. The easiest way to ensure this
is using devtools when in the package's directory:

```r
devtools::install_dev_deps()
```

# Adding a resource

## Checklist

- [ ] Add the new resource to `R/resources_backlog.R` following the standardized template
- [ ] Create a file for all necessary code to download your resource (`R/get_resource_name.R`)
- [ ] Include roxygen documentation for your resource following the provided template
- [ ] Check user-specified arguments (if any) for correctness
- [ ] Retrieve portfolio-wide parameters of interest for your resource from the portfolio
- [ ] Match the spatio-temporal extent of the portfolio with your resource
- [ ] Provide your own download functionality or use `.download_or_skip()`
- [ ] Delete any intermediate files that are no longer required
- [ ] Return the absolute file paths for all matching resource files
- [ ] Write a testthat script testing all the newly added functionality (except the actual download) and write it to `test/testthat/test-get_resource_name.R`
- [ ] Add a small example data set of your resource to `inst/res/resource_name/`
- [ ] Added a new dependency? Make sure to include a supporting statement for that dependency in your PR!

## Introducing a new resource to the backlog

A resource is a supported dataset that can be downloaded from a user's perspective  
via the <code>get_resources()</code> function. Currently, the package supports
only raster and vector resources. If you wish to submit the support of a new
resource, please be aware that we will only accept new resources if they are
associated with at least one indicator calculation. The very first step to
adding a resource is to add it into the internal resource backlog function so
that the package is aware of its existence. Once checkout to the new
branch and having the project opened in R Studio, issue the following
command to open the resource backlog file:

```{r open_resource_backlog, eval = FALSE}
file.edit("R/resources_backlog.R")
```

This file keeps track of all supported resources in a list object. You will
see that each resource shares a common structure and how it is specified. The name
of the list object will be the name the package uses to identify a specific resource.
Most importantly, the type argument specifies whether a resource is of type 'raster'
or 'vector'. If applicable, the source argument shall contain an URL pointing
to a webpage documenting the resource. The downloader argument is the package
internal function name that is used to download the resource. This function 
is the most important code file for a new resource. Then, arguments and their
default value to govern the download process can be specified. If no additional
arguments are needed just enter an empty list. For the `soilgrids` resource,
the internal backlog looks like this (don't spend too much effort in understanding
the arguments just yet. These will become clearer when we "write" the downloader.
When contributing a new resource, it is usually an iterative process between
the backlog and the downloader to 


y those arguments that need to be specified
by users):

```{r soilgrid_backlog, eval = FALSE}
soilgrids <- list(
  type = "raster",
  source = "https://www.isric.org/explore/soilgrids",
  downloader = ".get_soilgrids",
  arguments = list(
    layers = "clay",
    depths = "0-5cm",
    stats = "mean"
  )
)
```

With the resource being backlogged, the package now can find a 
resource called `soilgrids` of type `raster` and it can also identify the 
downloader function In this specific case, the package can also
determine the default values of three arguments in case users did not specify
anything. This is important information that will determine how the `get_resources()`
function works when called by users. 

## Documenting the new resource

By convention, the filename of a downloader **MUST** start with `get_<resource_name>.R` 
appended by the name of the resource. In the case of the soilgrids resource that
translates to `get_soilgrids.R`. In the first part of such a downloader, make
sure to include detailed documentation. This documentation should explain
what this resource represents, where it comes from (including a citation), 
and the arguments users should specify to control what is downloaded.
Importantly, this documentation **MUST** receive the roxygen tag `@docType data` as
well as the `@keywords resource` tag, so that the documentation can be 
identified as a resource. The `NULL` value below the documentation **MUST** be 
included. Below is a template that should be used for documenting a resource.

```r
#' Short title
#'
#' One or more description paragraphs might follow here. Please describe
#' required user arguments here, ideally as itemized lists.
#'
#' @name <the short name of your resource, same as in the backlog>
#' @docType data <we document resources as a dataset>
#' @keywords resource <identifies the documentation as a resource>
#' @format <one sentence on data format and spatial extent>
#' @references <ideally a citable scientific publication>
#' @source <a link in the \url{} tag linking to an online documentation>
NULL
```

## Function inputs for resources

After documenting the resource, you can get started with implementing the 
actual downloader. The downloader is a package's internal function that users do not
directly interact with. By convention, we append package internal function names
with a dot. Similar to the filename itself, resource downloaders should start
with `.get_<resource_name>`. The first argument is always `x`, which corresponds
to the portfolio object. Important attributes (e.g. the spatial-temporal extent)
can be derived from this object. Then additional user-facing arguments might follow.
After these arguments, each resource downloader receives the argument `rundir` which
by default should point to the output of `tempdir()`, but will be pointing to 
an output directory on disk where the output shall be written to when used by users.
Additionally, a logical called `verbose`, by default set to TRUE, controls 
the verbosity of the downloader as well as the dots argument. For the `soilgrids` 
resource, the function header thus looks like this:

```r
.get_soilgrids <- function(x, layers, depths, stats,
                           rundir = tempdir(),
                           verbose = TRUE,
                           ...) {
                           # downloader coder goes here
                           }
```

## Check arguments and retrieval of portfolio-wide parameters

Before actually conducting any downloads, it is important that you as the provider
of the new resource check extensively that all required arguments were correctly
specified. That specifically applies to the user-defined variables that your
downloader requires. The package framework cannot check
for the correctness of these arguments. That is something that each downloader
has to take care of. If some arguments are wrongly specified, the function
should fail (via `stop()`) and gracefully inform users which arguments where
misspecified and which values represent valid values. You can head
over to the soilgrids downloader (use `file.edit("R/get_soilgrids.R")`) and 
analyse the first few lines of the file (up to line ~130) to see how the
inputs are checked for the soilgrids resource.

Some portfolio-wide parameters that might be important to your specific downloader
can be determined by analysing the `x` portfolio attributes. Currently, the following
attributes with regard to a resource download are set, when users initialize their portfolio:

```r
  attr(x, "nitems") <- nrow(x)
  attr(x, "bbox") <- st_bbox(x)
  attr(x, "years") <- years
  attr(x, "aria_bin") <- aria_bin
```

Your resource downloader should take care that with these user-specified arguments
and the portfolio-wide parameters the files matching the spatio-temporal 
extent of the portfolio are downloaded.
These can be queried with the following syntax with the temporal extent of
the portfolio as an example:

```r
attributes(x)$years
```


## Using helper functions

If you construct several URLs and associated local filenames that you wish
to iterate over, the package provides a helper function to download these files
and skip already existing ones. The following code snippet shows how
to use that function:

```r
aria_bin <- attributes(x)$aria_bin
verbose <- attributes(x)$verbose

.download_or_skip(urls = source_urls, filenames = target_filenames,
                  verbose = verbose, stubbornnes = 6, check_existencs = TRUE,
                  aria_bin = aria_bin)

```
This function will attempt to download the specified URLs to the corresponding 
local filenames. URLs for which a corresponding filename already exists will be
skipped and information about this will be issued as a message if verbose is set 
to TRUE. The stubbornness controls the number of retries of failed downloads.
If check_existence is set to TRUE, `RCurl::url.exists(url)` will be used
to check if a given URL exists online. This check takes some time to
run, but for some resources, it can be useful if you cannot be sure if the constructed
URL exists on the remote location. Users are enabled to specify
an executable aria2c installation to support parallel downloads. The value of
the aria_bin variable will be `NULL` if no valid executable has been specified.
Otherwise, the downloader will use the aria2c program to download the specified URLs.

You are free and also encouraged to develop helper functions for your resource to 
increase the understandability and ease maintenance of your downloader. Any helper
functions associated with your resource should be located in the same file. If 
you feel a helper function might serve a purpose across different resources,
feel free to raise this in a comment and we can consider moving it to `R/utils.R`.
Since helper functions are internal to the package, they **MUST** start with a dot.
Package internal functions do not require a roxygen documentation. If you wish
to include documentation to make your code easier to understand, make sure to
add the `@keywords internal` and `@noRd` tags to your functions.

## Defining the output and handling intermediate files

You can create and delete intermediate files and directories within `rundir`.
The expected output of a resource downloader is a character vector
including the local file paths to all target files matching the spatio-temporal
extent of the portfolio `x`. For raster resources, a tile index indicating the
location on earth for each raster file and its file path will be constructed and
added as a resource to the portfolio. Vector resources are expected to be translated
to GeoPackages (`.gpkg`) and will be appended to the portfolio as is.

## Adding sample resource for package internal testing

We ask you to provide a small subset of your resource to `inst/res/resource_name`
so that indicators that depend on the resource can be tested without the need
to actual download the resource. Because there are some restrictions to the final
size of the package, we ask you to put substantial effort in reducing the size of
the files to a minimum. This includes cropping all resource samples to the spatial
extent of the polygon provided in `inst/extdata/sierra_de_neibe_478140.gpkg` or
a polygon of similar size supplied by you in case it the spatial extent does
not intersect with your resource. For raster resources, if the original raster
is encoded as float, consider changing the data type to integer by introducing a
scale factor. Also, please use a compression algorithm to further reduce the file 
size. For vector resources, consider reducing the number of vertices in case
the geometries are very complex.

## A note on dependencies for resources

Note, that a resource **SHALL NOT** add additional dependencies to the package. If
you add dependencies we require you to add a supporting statement to your PR
explaining why these dependencies are needed and why other approaches would fail.
Before accepting your PR, we might request you to change your code to remove 
these dependencies, if it is feasible to achieve the same functionality without.

# Adding an indicator

The process of adding an indicator is very similar to the one for resources.
However, some input-output requirements are actually different. Note, that in case
that you added a new resource we also expect a new indicator taking advantage of
that resource in your PR. 
As you will see, there are two new important concepts to have in mind when
adding an indicator. These are the processing mode and computational engines.
We will briefly explain these concepts below, however, you can also head over
to the [Terminology](terminology.html) vignette if you are interested in a more
comprehensive definition of these two terms.

## Checklist

- [ ] Add the new indicator to `R/indicators_backlog.R` following the standardized template
- [ ] Create a file for all necessary code compute your indicator (`R/calc_indicator_name.R`)
- [ ] Include roxygen documentation for your indicator following the provided template
- [ ] Check user-specified arguments (if any) for correctness
- [ ] Retrieve portfolio-wide parameters of interest for your indicator from the asset/portfolio
- [ ] Implement different computation engines for your indicator
- [ ] If applicable, implement both, asset and portfolio based processing modes
- [ ] Return a tibble in long format (no variables "hidden" in column names)
- [ ] Write a testthat script testing all the newly added functionality write it to `test/testthat/test-calc_indicator_name.R`, use snapshots to check for the correctness of numeric outputs
- [ ] Added a new dependency? Make sure to include a supporting statement for that dependency in your PR,
check if the dependency is installed, and add it to the Suggests field in the DESCRIPTION file


## Introducing a new indicator to the backlog

An indicator is a logical routine depending on one or more resources that
extracts numeric outputs for all assets in a portfolio. From a user's perspective,  
indicators are processed via the <code>calc_indicators()</code> function. 
We realized, that for large (potentially global) portfolios, depending on the
spatial resolution of a resource, different processing modes substantially
decrease the time needed for a computation. For high to medium resolution raster
resources, processing on the asset level benefits computation time. However,
spatially cropping coarse resolution datasets for a high number of assets introduces
significant overhead, thus processing these resources on a portfolio level
makes more sense.

You are asked to provide the most sensible approach to your indicator when you 
submit it to the internal backlog. When adding an indicator there, the package 
is aware made aware of its existence. You can issue the following
command to open the resource backlog file (or open it manually):

```{r open_indicator_backlog, eval = FALSE}
file.edit("R/indicator_backlog.R")
```

This file keeps track of all supported indicators in a list object. You will
see that each indicator shares a common structure. The name of the list object 
will be the name the package uses to identify a specific indicator. Note, 
that the name of an indicator **MUST NOT** be equal to any other indicator or resource.
The `processor` argument specifies the name of the function that you will
provide for the indicator calculation. The `inputs` argument refers to the 
supported resources that are required inputs for your indicator. As a requirement,
at least one resource needs to be specified, but your indicator can also depend
on more resources. We also ask you to provide the type of the resource (e.g. `raster` or `vector`).
In the `arguments` list object, you can specify any additional arguments that
users need to specify if they want to call your indicator function. If there
are no additional arguments, simply add an empty list. Please put sensible
default values in case your indicator function requires some arguments. Finally,
the `processing_mode` governs what your indicator functions will receive as inputs.
In case you set it to `asset`, each call to your indicator function will receive 
a single asset and the required resources crop to the spatial extent of that 
asset. If you set it to `portfolio` the function will receive the whole 
portfolio object and all required resource at the spatial extent of the portfolio.
Below, you will find an example for the `precipitation` indicator:


```{r precipitation_backlog, eval = FALSE}
precipitation <- list(
  processor = ".calc_precipitation",
  inputs = list(chirps = "raster"),
  arguments = list(
    scales_spi = NULL,
    engine = "extract"
  ),
  processing_mode = "portfolio"
)
```

If neither of the two processing modes lead to satisfactory processing times for 
your indicator, please leave an issue/comment to discuss the addition of 
another processing mode with the maintainers of the package. 

## Documenting the new indicator

By convention, the filename of a indicator processor **MUST** start with `calc_indicator_name.R` 
appended by the name of the resource. In the case of the precipitation indicator that
translates to `calc_precipitation.R`. In the first part of such a indicator processor, 
make sure to include detailed documentation. This documentation should explain
how it derives its numeric output, which resources are required for its calculation, 
and the arguments users should specify to control its functioning.
Importantly, this documentation **MUST** receive the roxygen tag `@docType data` as
well as the `@keywords indicator` tag, so that the documentation can be 
identified as an indicator The `NULL` value below the documentation **MUST** be 
included. Below is a template that should be used for documenting an indicator

```r
#' Short title
#'
#' One or more description paragraphs might follow here. Please describe
#' required resource and user arguments here, ideally as itemized lists.
#' Please document which processing engines are available for your indicator
#' and briefly describe how the indicator is derived from its inputs.
#'
#' @name <the short name of your indicator, same as in the backlog>
#' @docType data <we document indicators as a dataset>
#' @keywords indicator <identifies the documentation as an indicator>
#' @format <one sentence on number of columns, columns names of ouput tibble>
NULL
```


## Function inputs for indicators

After documenting the indicator, you can get started with implementing the 
actual processor. An indicator processor is a package's internal function that 
users do not directly interact with. By convention, we append package internal 
function names with a dot. Similar to the filename itself, indicator processors 
should start with `.calc_indicator_name>`. 
The first argument and indicator processor receives is always `shp`, which corresponds
to a single asset of the portfolio if you specified `processing_mode` as `"asset"` 
or the entire portfolio object if you specified `"portfolio"`. Important 
attributes (e.g. the spatial-temporal extent) can be derived from this object,
irrespective if it represents a single asset or the whole portfolio. Then,
one or more of the required resource are to be specified with the exact
names as they are included in the resource backlog. Again, in case the
`processing_mode` is set to `"asset"`, these resource will be spatial cropped
to the extent of the single asset, in case it is set to `"portfolio"` the
complete spatial extent of the portfolio is included.
After the resource, any user-defined additional arguments will follow. You
can assume that if users did not specify an argument, the default values
from the indicator backlog will be inserted instead. Note, that 
we define `engine` as an argument which is set by the users in order to give
them more fine-control how the output is computed. We will have a close look 
at engines in below.

All indicator functions also receive the argument `rundir`, where intermediate
files can be written to. You do not have to take care of cleaning that directory,
since the framework will clean up after the processing is done. Also, a logical
controlling the verbosity is handed over that you should use to decide
whether or not to print additional informative messages. For raster resources,
we included a logical `todisk` governing if intermediate raster files shall
be kept in memory or written to disk. Currently, the decision has to be 
handled by your indicator processor, however, we are evaluating possibilities
to determine the behaviour within the package itself. Last, the processing
mode is specified. This can be helpful if you wish to supply your processor
with the possibility to support either of the two modes. For the `precipitation`
indicator, the function header will look like this:

```r
.calc_precipitation <- function(shp,
                                chirps,
                                scales_spi = NULL,
                                engine = "extract",
                                rundir = tempdir(),
                                verbose = TRUE,
                                todisk = FALSE,
                                processing_mode = "portfolio",
                                ...) {
                                # processor logic goes here
                                }
```

## Check arguments and retrieval of portfolio-wide parameters

Before actually conducting any computation, it is important that you as the provider
of the new indicator check extensively that all required arguments were correctly
specified. That specifically applies to the user-defined variables that your
processor requires. The package framework cannot check
for the correctness of these arguments. If some arguments are wrongly specified, 
the function should fail (via `stop()`) and gracefully inform users which 
arguments where misspecified and which values represent valid values. 
You can head over to the precipitation processor (use `file.edit("R/calc_precipitation.R")`) and 
analyse the first few lines of the file (up to line ~76) to see how the
inputs are checked. Also note, that in case the required resource are `NULL` 
(that is the default value if an asset does not intersect with a resource), or
if any other configurations (e.g. years only smaller 1981 for the case of the
precipitation indicator) prevent a sensible processing of an indicator, we simply
return `NA`. The package will then fill in the tibble values for that asset 
with `NA`, so users now that the given indicator could not be calculated there. 

Some portfolio-wide parameters that might be important to your specific indicator
routine, and they can be derived via the `attributes()` function, e.g. for 
the years attribute:

```r
years = attributes(shp)$years
```

## Using helper functions

There are some package internal helper function that we found to be of use 
for multiple indicators that you are free to use in your indicator processor.
You will find them in `R/utils.R`. These helpers currently are:

- .check_available_years(): Checks if a given target year vector is available 
for a given indicator
- .check_engine(): Checks if a user-specified engine is available
- .check_stats(): Stats if a user-specified zonal statistic is available

You are encouraged to write your own helper function that are needed for your
indicator processor. These should be located in the same file as the main 
processor, start with a dot and should not be exported. If you wish to
include roxygen documentation for your helpers, make sure to
add the `@keywords internal` and `@noRd` tags to your functions. If you feel
that one or more of your helper functions would be of benefit to more that just 
one indicator, please comment in and issue/pull-request to discuss with the 
package maintainers if your helper function could be moved to `R/utils.R`.

## Adding engines to your indicator processor

In writing this package we realized that depending on the structure of a portfolio
(i.e. the number of assets, their size and geometric complexity), different engines
might lead to better processing times. We thus included three different engines for
most of our indicators, and we would invite you to do the same for your contribution.
Engines are mostly used in the very last step of an indicator calculation, that
is when some kind of zonal statistics are calculated for a specific asset.
The currently used engines are:

- terra::extract(): Takes a SpatRaster and a SpatVector as input and computes a
zonal statistic for all pixels within the SpatVector
- terra::zonal(): Takes two SpatRasters as input, one the with the target variable(s)
the other representing the rasterized input polygon. Then a zonal statistic
for the pixels that correspond to the asset extent is calculated
- exactextractr::exact_extract(): Takes a SpatRaster and an sf-object as input and
calculates a zonal statistic. It is implemented in C, thus promising fast processing
even for very large extents. 

If you wish to include another processing engine for your indicator, please indicate
this in a comment so that it can be discussed with the packages maintainers. 
Note, that indicators ideally should not add new dependencies if possible. If
they do, please add a supporting statement why this dependency is necessary for your
indicator. We also ask you to add dependencies to the `Suggests` field of the
`DESCRIPTION` file and that you check whether or not this dependency is installed
at the beginning of your indicator routine.


## Defining the output of indicator functions

Indicator functions, should return a tibble in long format as their output without
"hiding" any variables in column names. Beside from that requirement, the output
of your indicator does not need to follow any specific shape, except that columns
shall be equal across all assets. In case that you cannot calculate the indicator
for a specific indicator (e.g. because the extents do not overlap), simply
return `NA`. The package will handle these values internally and fill in `NA` 
a single row for that asset with the same column names as any other assets, its
values set to `NA`.


