---
title: "Contributing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contributing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

If you are reading this vignette you are most probably to contribute to the `{mapme.biodiversity}`
package. This is great news and we are very happy to receive Pull-Requests
extending the package's functionality! Below you will find important in-depth
information about how to add resources and indicators to make the process
as seamless as possible for both you and the package's maintainers. Please make
sure to read and understand this guide before opening a PR. If in doubt, especially
if you feel that the framework does not support your use case, always
feel free to raise an issue and we will happily discuss how we can support your
ideas. If you have not already done so, make sure to read the [Terminology](terminology.html)
vignette to get familiar with the most important concepts of this package.

Note that we use the [tidyverse style guide](https://style.tidyverse.org/) 
for the package. That specifically means that function and variable names should 
follow the snake case pattern. We also use the arrow assignment operator (`<-`). 
When submitting a PR that does not consistently follow the tidyverse style guide, 
the maintainers of the package might change the code to adhere to this code style
without further notice before accepting the PR. 

# Getting started

Ideally, you clone the GitHub repository via the git command in a command line
on Linux and MacOS systems or via the GitHub Desktop application on Windows.
On Linux, the command would look like this:

```bash
git clone https://github.com/mapme-initiative/mapme.biodiversity
```

We do not accept pushes to main, thus the first step would be to create
a specific branch for your extension. In this tutorial, we will pretend to 
re-implement the `soilgrids` resource and the associated `soilproperties` 
indicator, so we will create a branch reflecting this. Don't forget to check out
to the newly created branch!

```bash
git branch add-soilgrid-indicators
git checkout add-soilgrid-indicators
```
Below, we will assume that you develop your extension to the package in R Studio.
The general guidelines to follow also apply if you choose different tooling 
for your development process, however, it will not be covered in this vignette.
We assume that all R development dependencies are installed. The easiest way to 
ensure this is using `{devtools}`:

```r
devtools::install_dev_deps()
```

# Adding a resource

## Checklist

- [ ] Create a file for all necessary code to download your resource (`R/get_<resource_name>.R`)
- [ ] Include roxygen documentation for your resource following the provided template
- [ ] Create outer-level function for user facing arguments
- [ ] Check user-specified arguments (if any) for correctness
- [ ] Create inner-level function with standard arguments
- [ ] Match the spatio-temporal extent of the portfolio with your resource
- [ ] Provide your own download functionality or use `download_or_skip()`
- [ ] Delete any intermediate files that are no longer required
- [ ] Return the absolute file paths for all matching resource files
- [ ] Write a testthat script testing all the newly added functionality (except the actual download) and write it to `test/testthat/test-get_<resource_name>.R`
- [ ] Add a small example data set of your resource to `inst/res/<resource_name>`
- [ ] Add useful information about the resource to the register via `register_resource()`
- [ ] Added a new dependency? Make sure to include a supporting statement for that dependency in your PR!

## Overview of adding a resource

A resource is a supported dataset that can be downloaded from a user's perspective
by specifying one or more functions to `get_resources()`. Currently, the package 
supports only raster and vector resources. If you wish to submit the support of 
a new resource, please be aware that we will only accept new resources if they are
associated with at least one indicator calculation. The very first step to
adding a resource is to create a new file that will be holding the required code.  
Once you checked out to the new branch and having the project opened in R Studio, 
adapt the following command to open the a new resource file:

```{r open_resource_file, eval = FALSE}
file.edit("R/get_<your-new-resource>.R")
# e.g. file.edit("R/get_soildgrids.R")
```

## Documenting the new resource

In the first part of a resource function, make sure to include detailed 
documentation. This documentation should explain what this resource represents, 
where it comes from (including a citation), and the user-facging arguments that 
should be specified during runtime. 
Importantly, this documentation **MUST** receive the roxygen tag 
`@keywords resource`, so that the documentation will be identified as a resource. 
Also, add the bare name of the resource as the `@name` tag (e.g. in the case
of our example that translates to `@name soilgrids`).

```r
#' Short title
#'
#' One or more description paragraphs might follow here. Please describe
#' the spatio-temporal structure of your resource here briefly.
#'
#' @name <the short name of your resource>
#' @param <any user-facing arguments>
#' @keywords resource <identifies the documentation as a resource>
#' @references <ideally a citable scientific publication>
#' @source <a link in the \url{} tag linking to an online documentation>
#' @returns A function that makes a resource available for a portfolio
#' @include register.R
#' @export
```

The last two tags are important to add as well. The include statement is 
mandatory for the register functionality (more on that below) to be loaded
before your resource function. The export tag is important so that the resource
is actually exposed to the users of the package.

## Constructing a resource function - Outer level

Resource functions are constructed as closures, i.e. functions that return a function.
The outer level exposes arguments to be set by users of the function to fine-control
the flow of the function. Note, it is important to check the user input in this
outer level for correctness so that warning/error messages in case of any
miss specifications are thrown immediately.

For soilgrids, this outer level could look something like this:

```r
get_soilgrids <- function(layers, depths, stats) {

  stopifnot(all(!missing(layers), !missing(depths), !missing(stats)))
  stopifnot(all(layers %in% c("clay", "sand", "silt")))
  stopifnot(all(depths %in% c("0-5cm", "5-15cm", "15-30cm")))
  stopifnot(all(stats %in% c("Q0.05", "Q0.5", "mean", "Q0.95")))

# .... inner function level

}

```

There are some exported helper functions for re-occurring argument checks that
you are free to use (e.g. `check_available_years()`).
Note, that the arguments defined this way in the outer level
of the resource function are then ready to be used in the inner level that we
will have a look at next.

## Constructing a resource function - Inner level

The inner level of a resource function has a mandatory function signature that
will be checked during run-time. Your function is required to exactly specify 
the below signature. For the soilgrids resource, this looks like this:

```r 
  function(x,
           name = "soilgrids",
           type = "raster",
           outdir = mapme_options()[["outdir"]],
           verbose = mapme_options()[["verbose"]],
           testing = mapme_options()[["testing"]]) {
           
           # ... function body
}
```

The `x` argument here represents the portfolio object handed over by the user 
when calling `get_resources()` which is an `sf`-object and can thus be used 
to derive the spatial extent of the portfolio. Next, comes the name and the 
type of the resource which is required for the backend to correctly handle the 
output and log the resource once it has been made available. 

The other arguments should default to the their respective output values of 
`mapme_options()` and represent a character vector for the output directory, 
a logical to control the verbosity and another logical indicating if the code 
is currently executed in a testing mode. We will look into how these things come 
together now as we peak into constructing the actual body of a resource function.

## Constructing a resource function - Body

The expected output of a resource function is a character vector of file paths
to either raster or vector sources that represent GDAL readable spatial data sets
on the local file system. In case you require to download or write intermediate
files, you can create files in the output of `tempdir()`. 

The output files should match the spatial extent of the portfolio and respect
the arguments specified by the user. Output files should be constructed based on the
`outdir` argument. The function should not re-download already existing
files in the output directory. For flat files you can use the `download_or_skip()`
helper function to make sure that existing files are not re-downloaded.

Use the `verbose` argument to decide if informative messages should be printed,
e.g. to inform users about download progress. Errors or warnings should be emitted
in either case. 

Please include a check if `testing = TRUE` that returns example filenames early
without actually downloading anything. This is mandatory for automated test checks
of the package on CI platforms where it is not possible to conduct lengthy downloads.

If there is now intersection between the `x` object and your resource, make sure
to return `NA` as early as possible. 

## Adding sample resource for package internal testing

We ask you to provide a small subset of your resource to `inst/res/resource_name`
so that indicators that depend on the resource can be tested without the need
to actual download the resource. 

Because there are some restrictions to the final size of the package, we ask you 
to put substantial effort in reducing the size of the files to a minimum. 
This includes cropping all resource samples to the spatial extent of the polygon 
provided in `inst/extdata/sierra_de_neibe_478140.gpkg` or a polygon of similar 
size supplied by you in case it the spatial extent does not intersect with your 
resource. 

For raster resources, if the original raster is encoded as float, consider 
changing the data type to integer by introducing a scale factor. Also, please 
use a compression algorithm to further reduce the file size. For vector resources, 
consider reducing the number of vertices in case the geometries are very complex. 

Finally, put your processing script of the resource into `data-raw` to 
ensure reproducibility. Then, you are required to write a unit-test for
your resource function, which should execute as much of your code as possible
without actually conducting a download. 

## A note on dependencies for resources

Note, that a resource **SHALL NOT** add additional dependencies to the package. If
you add dependencies we require you to add a supporting statement to your PR
explaining why these dependencies are needed and why other approaches would fail.
Before accepting your PR, we might request you to change your code to remove 
these dependencies, if it is feasible to achieve the same functionality without.

# Adding an indicator

The process of adding an indicator is very similar to the one for resources.
However, some input-output requirements are different. Note, that in case
that you added a new resource we also expect a new indicator taking advantage of
that resource in your PR. 

As you will see, there are two new important concepts to have in mind when
adding an indicator. These are the processing mode and computational engines.
We will briefly explain these concepts below, however, you can also head over
to the [Terminology](terminology.html) vignette if you are interested in a more
comprehensive definition of these two terms.

## Checklist

- [ ] Create a file for all necessary code compute your indicator (`R/calc_<indicator_name>.R`)
- [ ] Create outer-level function for user facing arguments
- [ ] Check user-specified arguments (if any) for correctness
- [ ] Create inner-level function with standard arguments
- [ ] If applicable, implement both, asset and portfolio based processing modes
- [ ] Return a tibble in long format (no variables "hidden" in column names)
- [ ] Write a testthat script testing all the newly added functionality write it to `test/testthat/test-calc_<indicator_name>.R`
- [ ] Added a new dependency? Make sure to include a supporting statement for that dependency in your PR!

## Overview of adding a new indicator

An indicator is a logical routine depending on one or more resources that
extracts numeric outputs for all assets in a portfolio. From a user's 
perspective, indicators are processed via the `calc_indicators()` function.
You as a developer will have to construct an indicator function as a 
closure, e.g. a function that returns another function. The outer level
exposes user-facing arguments and checks that they are correctly specified,
while the inner level is required to follow a specified signature and returns
a tibble.

Once you checked out to the new branch and having the project opened in R Studio, 
adapt the following command to open the a new indicator file:

```{r open_indicator_file, eval = FALSE}
file.edit("R/calc_<your-new-indicator>.R")
# e.g. file.edit("R/calc_precipitation")
```

## Documenting the new indicator

In the first part of an indicator function, make sure to include detailed 
documentation. This documentation should explain which resources are required
to calculate the indicator, the user-facing arguments that 
should be specified during runtime and the structure of the output tibble. 
Importantly, this documentation **MUST** receive the roxygen tag 
`@keywords indicator`, so that the documentation will be identified as an indicator. 
Also, add the bare name of the indicator as the `@name` tag (e.g. `@name <indicator>`).

```r
#' Short title
#'
#' One or more description paragraphs might follow here. Please describe
#' required resource and user arguments here.
#' Please document which processing engines are available for your indicator
#' and briefly describe how the indicator is derived from its inputs.
#'
#' @name <the short name of your indicator, same as in the backlog>
#' @param <any user-facing arguments>
#' @keywords indicator <identifies the documentation as an indicator>
#' @returns A function that calculates an indicator for a portfolio
#' @include register.R
#' @export
```

The last two tags are important to add as well. The include statement is 
mandatory for the register functionality (more on that below) to be loaded
before your indicator function. The export tag is important so that the resource
is actually exposed to the users of the package.


## Constructing an indicator function - Outer level

Indicator functions are constructed as closures, i.e. functions that return a function.
The outer level exposes arguments to be set by users of the function to fine-control
the flow of the function. Note, it is important to check the user input in this
outer level for correctness so that warning/error messages in case of any
miss specifications are thrown immediately.

For `elevation`, this outer level could look something like this:

```r
calc_elevation <- function(engine = "extract",
                           stats = "mean") {
                           
  engine <- check_engine(engine)
  stats <- check_stats(stats)
  
  # ... inner function level
  
}

```

There are some exported helper functions for re-occurring argument checks that
you are free to use (e.g. `check_engine()`). Note, that the arguments defined 
this way in the outer level of the indicator function are then ready to be used 
in the inner level that we will have a look at next.

## Constructing an indicator function - Inner level

The inner level of an indicator function has a mandatory function signature that
will be checked during run-time. Your function is required to exactly specify 
the below signature. For the `elevation` indicator, this looks like this:

```r 
  function(x,
           nasa_srtm = NULL,
           name = "elevation",
           mode = "asset",
           verbose = mapme_options()[["verbose"]]) {
           
           # ... function body
}
```

The `x` argument here represents the portfolio object handed over by the user 
when calling `get_resources()` which is an `sf`-object with `'POLYGON'` as features.
Next, comes the name(s) of the required resource(s) and the name of the indicator.
What follows is the computation mode, that must be one of `"asset"` or `"portfolio"`.

We realized, that for large (potentially global) portfolios, depending on the
spatial resolution of a resource, different processing modes substantially
impact the time needed for a computation. For high to medium resolution raster
resources, processing on the asset level benefits computation time. However,
spatially cropping coarse resolution datasets for a high number of assets
introduces significant overhead, thus processing these resources on a 
portfolio level is more efficient.

If neither of the two processing modes lead to satisfactory processing times for 
your indicator, please leave an issue/comment to discuss the addition of 
another processing mode with the maintainers of the package. 

The argument `verbose` defaults to the corresponding package-wide option and 
should control the verbosity of you indicator function.

## Constructing an indicator function - Body

The expected output of an indicator function is a tibble. Depending on the 
`mode` specified for processing, it is a single tibble for `mode = "asset"`,
or a list of tibbles equal to the rows of `x` in case `mode = "portfolio"`.

You may use helper functions provided by the package for a common interface e.g.
for vector-raster zonal statistics (e.g. by using `select_engine()`).

You are encouraged to write your own helper function that are needed for your
indicator processor. These should be located in the same file as the main 
processor, start with a dot and should not be exported. 

If you wish to include roxygen documentation for your helpers, make sure to
add the `@keywords internal` and `@noRd` tags to your functions. If you feel
that one or more of your helper functions would be of benefit to more that just 
one indicator, please comment in and issue/pull-request to discuss with the 
package maintainers if your helper function could be moved to `R/utils.R`.

Use the `verbose` argument to decide if informative messages should be printed,
e.g. to inform users about processing progress. Errors or warnings should be emitted
in either case. 

If there is no intersection between the `x` object and the required resources,
or for any other reason why your indicator might not be calculated with the
given configuration, make sure to return `NA` as early as possible. 


## Adding units tests for an indicator

You are required to add unit tests for your indicator using the package internal
example data sets for resources. Make sure to properly test for missspecification
of user-facing arguments and also check for the correctness of numerical results
of your indicator.


