---
author: "Darius A. Görgen"
title: "Installation"
date: "Last modified: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This sections describes how to install the `{mapme.biodiversity}` package. 
Because the package handles some spatial data, we assume that you have
a working installation of [GDAL](https://gdal.org/index.html) and [PROJ](https://proj.org/) 
on your system. The GDAL installation needs to be found on your PATH variable, 
because several system calls will be issued when the package is used. Please 
refer to the respective installation guides to learn about how to install these 
programs on your OS.
While the package supports sequentially downloading all supported resources,
we include the option to provide a path to an [aria2](https://aria2.github.io/) executable 
in order to seed up the download based on parallel processing. We expect that 
the path to the executable binary is known to the user and we test it if 
provided during portfolio initialization.

Provided that above requirements are met, installing the `{mapme.biodiversity}` 
package is then achieved trough the following function call: 


```{r installation, eval = F}
remotes::install_github("mapme-initiative/mapme.biodiversity")
```

The installation process will install potential missing R dependencies.
After the successful installation, you are ready to use the package's 
functionality. Simply load the package for your R session with the library
call:

```{r setup, val = F}
library(mapme.biodiversity)
```

