---
title: "Installation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This section describes how to install the `{mapme.biodiversity}` package. 
Since the package handles spatial data, we assume that you have
a working installation of [GDAL](https://gdal.org/index.html) and [PROJ](https://proj.org/) 
on your system. Our recommendation is to follow the `{sf}` installation instructions.

While the package supports sequentially downloading all supported resources,
we include the option to provide a path to an [aria2](https://aria2.github.io/) executable 
to speed up downloads based on parallel processing. We expect that 
the path to the executable binary is known to the user and we test if it is acually executable.

Given that the above requirements are met, installing the `{mapme.biodiversity}` 
current CRAN release is then achieved through the following function call: 

```{r installation-cran, eval = FALSE}
install.packages("mapme.biodiversity")
```

You can install the latest development version from GitHub via the following
call:

```{r installation, eval = FALSE}
remotes::install_github("mapme-initiative/mapme.biodiversity")
```

The installation process will install potential missing R dependencies.
After the successful installation, you are ready to use the package's 
functionality. Simply load the package for your R session with the library
call:

```{r setup, eval = FALSE}
library(mapme.biodiversity)
```

