---
title: "How To: Transform indicator output to wide-format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How To: Transform indicator output to wide-format}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sf)
library(dplyr)
library(tidyr)
library(mapme.biodiversity)
```

## Objectives

This tutorial gives you some information how to transform the output of
the mapme-biodiversity package to a wide format for exchange with other
(geospatial-)software, such as QGIS. This is necessary because the
package uses the so-called nested-list format by default to support very
different indicators that each may come in its own format. However, this
format is specific to R and to use the data in other software thus
requires some additional steps to be taken. This vignette thus shows you
how you can obtain a wide-format for your portfolio that you can easily
export and use with other software.

## What are long vs. wide tables?

Tabular data can be structured in two different ways, which are usually
referred to as long and wide format. Most people are more familiar with
the wide format, because this is a format we as humans would naturally
structure our data when we work with spreadsheets, e.g. in Excel. In
wide-format, the identifier of a observation is included exactly once
and does not repeat itself (see Table A). In the long format, the
identifier as well as other qualifying variables, might be repeated
several times to uniquely identify each observation is a single row (see
Table B). The long format is often required when interacting with
computers, e.g. to make plots with ggplot2. The content of the two is
exactly the same either way, the one might be just more friendly to
humans than to computers. If you are familiar with the R tidyverse, you
might also have heard of the term *tidy* data. In terms of tabular data
you can imagine *tidy* data to be referring to data in a long table
which naturally fulfills the following requirements:

-   each variable has its own column
-   each observation has its own row
-   each value has its own cell

![Fig. 1: Example of a wide and long table holding the same
data.](assets/img/wide-long-tables.png)

Table A, in that sense, is not *tidy* since the year variable is not
found in its own column but instead it is scattered in two different
columns. Table B is a long format with each variable being found in
exactly one column. In that sense, each individual row represents
exactly one observation, meaning the observation of a specific country
in a specific year.

When we structure data in the long format the objects will usually have
a larger memory footprint than compared to a wide format. For smaller
objects or data types with small memory consumption, this might not pose
a serious limitation to the workflow. However, geometry information,
here indicated as a WKT string, might quickly accumulate a large
proportion of the available memory, even more so if the portfolio
consists of a high number of complex geometries that are then copied to
fit the the long-format requirement. For that reason, this packages uses a 
nested-list format to hold tables for indicators as single columns within the 
portfolio. The remainder of this tutorial will show you in more detail
how you can work with those R specific data format.

## The simple case - single-row indicators

We start by reading a GeoPackage from disk. For the sake of the
argument, we split the original single polygon into 9 distinct polygons
to simulate a realistic portfolio consisting of multiple assets.

```{r read-poly}
aoi <- read_sf(
  system.file("extdata", "sierra_de_neiba_478140_2.gpkg",
    package = "mapme.biodiversity"
  )
)
aoi <- st_as_sf(st_make_grid(aoi, n = 3))
print(aoi)
```

As a simple example, suppose we interested in extracting the average traveltime
to cities between 20,000 and 50,000 inhabitants in our portfolio. As usual, 
we will have to make available the Nelson et al. resource as well as requesting 
the calculation of the respective indicator.

```{r calc-elev}
mapme_options(
  outdir = system.file("res", package = "mapme.biodiversity"),
  verbose = FALSE
)

port_w_nelson <- get_resources(aoi, get_nelson_et_al(ranges = "20k_50k"))
port_w_traveltime <- calc_indicators(port_w_nelson, calc_traveltime(stats = "mean"))
print(port_w_traveltime)
```

As we can observe from the output, a new column was added to the `sf`
object. It is called `traveltime` and is of type `list` indicating 
that it represents a nested-list column. What that means is that we 
are able to maintain the rectangular shape of our original data 
(e.g. one polygon per row), while supporting arbitrarily shaped 
outputs for indicators. Let's observe how the traveltime indicator
looks like in this instance:

```{r investigate-elev}
print(port_w_traveltime$traveltime[[1]])
```

From the syntax above, you can see how you can access a single object
within the nested list column (e.g. by using the list accessor `[[`). In
this case, the shape of the traveltime indicator is a single-row and
two-column tibble with the average minutes and the distance category as its value. 
We can now use either of two functions to transform the portfolio to long
or wide formats:

```{r elev-to-wide}
portfolio_long(port_w_traveltime)
portfolio_wide(port_w_traveltime)
```

In either case, the result still has 9 rows just like the original data frame.
In this particular scenario, where the is only a single variable for each asset, 
the unnested portfolio contains just one value per asset.

We could serialize the object to disk in either format by calling `write_portfolio()`:

```{r simple-export}
dsn_long <- tempfile(fileext = ".gpkg")
dsn_wide <- tempfile(fileext = ".gpkg")
write_portfolio(port_w_traveltime, dsn_long, format = "long")
write_portfolio(port_w_traveltime, dsn_wide, format = "wide")
```

## The harder case - indicators with multi-row output

Let's continue to query an indicator which has a multi-row output, i.e.
precipitation statistics from [WorldClim](https://www.worldclim.org/).

```{r calc-prec}
port_w_wc <- get_resources(aoi, get_worldclim_precipitation(years = 2010:2018))
port_w_prec <- calc_indicators(port_w_wc, calc_precipitation_wc(stats = "mean"))
print(port_w_prec)
```

We see that similar to the traveltime indicator, the output consists of 9
rows with the precipitation indicator as a single nested-list column.
Note, however, the differences when we take a look at the shape of the
indicator for a specific asset:

```{r investigate-precip}
print(port_w_prec$precipitation_wc[[1]])
```

For a single asset, we obtain a tibble with 12 rows (for each month in
the queried year 2018). Now, let's have a look at what happens if we transform
this table to long format:

```{r prec-unnest}
portfolio_long(port_w_prec)
```

Instead of 9 rows, we get a tibble with 108 rows (9 assets \* 12), with
the metadata for each asset as well as the geometry column being
repeated. For very large portfolios, this data layout might be very memory 
intensive. In these cases it is more favorable to transform the portfolio
to a wide layout.

```{r fun-prec-to-wide}
portfolio_wide(port_w_prec)
```

From the example output above we see that in this case, we obtain a resulting
object with 9 rows only. The indicator data is now found in the respective
columns which are named according to this schema:

`<indicator-name>_<datetime>_<variable>_<unit>`

with the values being found in the rows for each unique combination of this 
pattern.

Similar to the simple case, we can use `write_portfolio()` to serialize 
the portfolio object to disk in the desired layout:

```{r write-output}
dsn_long <- tempfile(fileext = ".gpkg")
dsn_wide <- tempfile(fileext = ".gpkg")
write_portfolio(port_w_prec, dsn_long, format = "long")
write_portfolio(port_w_prec, dsn_wide, format = "wide")
```
