---
title: "How To: Transform indicator output to wide-format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How To: Transform indicator output to wide-format}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(sf)
library(dplyr)
library(tidyr)
library(mapme.biodiversity)
```

## Objectives

This tutorial gives you some information how to transform the output of
the mapme-biodiversity package to a wide format for exchange with other
(geospatial-)software, such as QGIS. This is necessary because the
package uses the so-called nested-list format by default to support very
different indicators that each may come in its own format. However, this
format is specific to R and to use the data in other software thus
requires some additional steps to be taken. This vignette thus shows you
how you can obtain a wide-format for your portfolio that you can easily
export and use with other software.

## What are long vs. wide tables?

Tabular data can be structured in two different ways, which are usually
referred to as long and wide format. Most people are more familiar with
the wide format, because this is a format we as humans would naturally
structure our data when we work with spreadsheets, e.g. in Excel. In
wide-format, the identifier of a observation is included exactly once
and does not repeat itself (see Table A). In the long format, the
identifier as well as other qualifying variables, might be repeated
several times to uniquely identify each observation is a single row (see
Table B). The long format is often required when interacting with
computers, e.g. to make plots with ggplot2. The content of the two is
exactly the same either way, the one might be just more friendly to
humans than to computers. If you are familiar with the R tidyverse, you
might also have heard of the term *tidy* data. In terms of tabular data
you can imagine *tidy* data to be referring to data in a long table
which naturally fulfills the following requirements:

-   each variable has its own column
-   each observation has its own row
-   each value has its own cell

![Fig. 1: Example of a wide and long table holding the same
data.](assets/img/wide-long-tables.png)

Table A, in that sense, is not *tidy* since the year variable is not
found in its own column but instead it is scattered in two different
columns. Table B is a long format with each variable being found in
exactly one column. In that sense, each individual row represents
exactly one observation, meaning the observation of a specific country
in a specific year.

When we structure data in the long format the objects will usually have
a larger memory footprint than compared to a wide format. For smaller
objects or data types with small memory consumption, this might not pose
a serious limitation to the workflow. However, geometry information,
here indicated as a WKT string, might quickly accumulate a large
proportion of the available memory, even more so if the portfolio
consists of a high number of complex geometries that are then copied to
fit the the long-format requirement. For that\
reason, this packages uses a nested-list format to hold tables for
indicators as single columns within the portfolio. The remainder of this
tutorial will show you in more detail\
how you can work with those R specific data format.

## The simple case - single-row indicators

We start by reading a GeoPackage from disk. For the sake of the
argument, we split the original single polygon into 9 distinct polygons
to simulate a realistic portfolio consisting of multiple assets.

```{r read-poly}
aoi <- read_sf(
  system.file("extdata", "sierra_de_neiba_478140_2.gpkg",
    package = "mapme.biodiversity"
  )
)
aoi <- st_as_sf(st_make_grid(aoi, n = 3))
print(aoi)
```

As a simple example, suppose we interested in extracting the average
elevation per asset in our portfolio. As usual, we will have to make
available the SRTM resource as well as requesting the calculation of the
respective indicator.

```{r calc-elev}
port <- init_portfolio(
  aoi,
  years = 2018,
  outdir = system.file("res", package = "mapme.biodiversity"),
  tmpdir = system.file("tmp", package = "mapme.biodiversity"),
  add_resources = TRUE
)

port_w_srtm <- get_resources(port, "nasa_srtm")
port_w_elev <- calc_indicators(port_w_srtm, "elevation", stats_elevation = "mean")
print(port_w_elev)
```

As we can observe from the output, two new columns were added to the sf
object. The first is called `assetid` and simply is a unique identifier
of the respective asset. The second column is called `elevation` and is
of type `list` indicating that it represents a nested-list column. What
that means is that we are able to maintain the rectangular shape of our
original data (e.g. one polygon per row), while supporting arbitrarily
shaped outputs for indicators. Let's observe how the elevation indicator
looks like:

```{r investigate-elev}
print(port_w_elev$elevation[[1]])
```

From the syntax above, you can see how you can access a single object
within the nested list column (e.g. by using the list accessor `[[`). In
this case, the shape of elevation indicator is a single-row and
single-column tibble with the mean elevation as its value. Because the
indicator elevation comes without a temporal axis, we can simply unnest
the elevation column to get to a wide-format output:

```{r elev-to-wide}
port_wide_1 <- tidyr::unnest(port_w_elev, elevation)
print(port_wide_1)
```

In this case, the result has 9 rows just like the original data frame.
We can now export our portfolio for example as a geopackage that can be
read directly into e.g. QGIS or as a CSV without the geospatial
information:

```{r simple-export}
st_write(port_wide_1, dsn = tempfile(fileext = ".gpkg"))
write.csv(st_drop_geometry(port_wide_1), file = tempfile(fileext = ".csv"))
```

## The harder case - indicators with multi-row output

Let's continue to query an indicator which has a multi-row output, i.e.
precipitation statistics from [WorldClim](https://www.worldclim.org/).

```{r calc-prec}
port_w_chirps <- get_resources(port, "worldclim_precipitation")
port_w_prec <- calc_indicators(port_w_chirps, "precipitation_wc", stats_worldclim = "mean")
print(port_w_prec)
```

We see that similar to the elevation indicator, the output consists of 9
rows with the precipitation indicator as a single nested-list column.
Note, however, the differences when we take a look at the shape of a the
indicator for a specific asset:

```{r investigate-precip}
print(port_w_prec$precipitation_wc[[1]])
```

For a single asset, we obtain a tibble with 12 rows (for each month in
the queried year 2018). Now, if we just simply unnest this indicator
column observe what happens to the shape of the output:

```{r prec-unnest}
port_wide_2 <- tidyr::unnest(port_w_prec, precipitation_wc)
print(port_wide_2)
```

Instead of 9 rows, we get a tibble with 108 rows (9 assets \* 12), with
the metadata for each asset as well as the geometry column being
repeated. This is not desirable, especially for portfolios with large
numbers of assets where the repetition of the geometry data can cause
very large objects. Now, let's investigate how we still are able to get
a wide output. A valid approach is to re-shape the indicators per asset
to a wide format before unnesting the indicator column. For this, let's
write a function which takes the precipitation indicator of a single
asset as input and transforms it to a wide-format.

```{r fun-prec-to-wide}
prec2wide <- function(data) {
  pivot_wider(data, names_from = date, values_from = prec_mean, names_prefix = "prec-")
}
prec2wide(port_w_prec$precipitation_wc[[1]])
```

From the example output above we see that our custom `prec2wide()`
functions takes a single tibble as input, puts in the dates appended by
`prec-` as column names and the associated average precipitation values
as a single row. Now, we can use functionality of the `purrr` package to
map this function over our indicator column, save the result in a new
column (or simply overwrite the original), and finally simply call
unnest to get the output in a wide format.

```{r prec-to-wide}
port_w_prec$precipitation_wc2 <- purrr::map(port_w_prec$precipitation_wc, prec2wide)
port_w_prec <- select(port_w_prec, assetid, precipitation_wc2)
port_wide_3 <- unnest(port_w_prec, precipitation_wc2)

print(port_wide_3)
```

We can now use `st_write()` to save this wide-format sf object to a
geospatial format of our choice or drop the geometry information
altogether and dump the data into a csv.

```{r write-output}
st_write(port_wide_3, dsn = tempfile(fileext = ".gpkg"))
write.csv(st_drop_geometry(port_wide_3), file = tempfile(fileext = ".csv"))
```
