---
title: "How To: Transform indicator output to wide-format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How To: Transform indicator output to wide-format}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE}
library(sf)
library(dplyr)
library(tidyr)
library(mapme.biodiversity)
mapme_options(verbose = FALSE)
```

## Objectives

This tutorial gives you some information how to transform the output of
the mapme-biodiversity package to a wide format for exchange with other
(geospatial-)software, such as QGIS. This is necessary because the
package uses the so-called nested-list format by default to represent indicators. 
However, this format is specific to R and to use the data in other software thus
requires some additional steps to be taken. This vignette shows you
how you can change the data layout of your portfolio that you can easily
serialize to a spatial format of your choice and use with other software.

## What are long vs. wide tables?

Tabular data can be structured in two different ways, which are usually
referred to as long and wide format. Most people are more familiar with
the wide format, because this is a format we as humans would naturally
structure our data when we work with spreadsheets, e.g. in Excel. In
wide-format, the identifier of a observation is included exactly once
and does not repeat itself (see Table A). In the long format, the
identifier as well as other qualifying variables, might be repeated
several times to uniquely identify each observation in a single row (see
Table B). The long format is often required when interacting with
computers, e.g. to make plots with ggplot2. The content of the two is
exactly the same either way, the one might be just more friendly to
humans than to computers. If you are familiar with the R tidyverse, you
might also have heard of the term *tidy* data. In terms of tabular data
you can imagine *tidy* data to be referring to data in a long table
which naturally fulfils the following requirements:

-   each variable has its own column
-   each observation has its own row
-   each value has its own cell

![Fig. 1: Example of a wide and long table holding the same
data.](assets/img/wide-long-tables.png)

Table A, in that sense, is not *tidy* since the year variable is not
found in its own column but instead it is scattered in two different
columns. Table B is a long format with each variable being found in
exactly one column. In that sense, each individual row represents
exactly one observation, meaning the observation of a specific country
in a specific year.

When we structure data in the long format the objects will usually have
a larger memory footprint than compared to a wide format. For smaller
objects or data types with small memory consumption, this might not pose
a serious limitation to the workflow. However, geometry information,
here indicated as a WKT string, might quickly accumulate a large
proportion of the available memory, even more so if the portfolio
consists of a high number of complex geometries that are then copied to
fit the the long-format requirement. For that reason, this packages uses a 
nested-list format to hold tables for indicators as single columns within the 
portfolio. The remainder of this tutorial will show you in more detail
how you can work with those R specific data format.

## The simple case - single-row indicators

We start by reading a GeoPackage from disk. For the sake of the
argument, we split the original single polygon into 9 distinct polygons
to simulate a realistic portfolio consisting of multiple assets.

```{r read-poly}
aoi <- read_sf(
  system.file("extdata", "sierra_de_neiba_478140_2.gpkg",
    package = "mapme.biodiversity"
  )
)
aoi <- st_as_sf(st_make_grid(aoi, n = 3))
print(aoi)
```

As a simple example, suppose we interested in extracting the average traveltime
to cities between 20,000 and 50,000 inhabitants in our portfolio. As usual, 
we will have to make available the Nelson et al. resource as well as requesting 
the calculation of the respective indicator.


```{r init-data-dir, echo=FALSE}
outdir <- file.path(tempdir(), "mapme-resources")
mapme.biodiversity:::.copy_resource_dir(outdir)
```


```{r calc-elev}
outdir <- file.path(tempdir(), "mapme-resources")
dir.create(outdir, showWarnings = FALSE)

mapme_options(
  outdir = outdir,
  verbose = FALSE
)

aoi <- get_resources(aoi, get_nelson_et_al(ranges = "100k_200k"))
aoi <- calc_indicators(aoi, calc_traveltime(stats = "mean"))
print(aoi)
```

As we can observe from the output, a new column was added to the `sf`
object. It is called `traveltime` and is of type `list` indicating 
that it represents a nested-list column. What that means is that we 
are able to maintain the rectangular shape of our original data 
(e.g. one polygon per row), while supporting arbitrarily shaped 
outputs for indicators. Let's observe how the `traveltime` indicator
looks like in this instance:

```{r investigate-elev}
print(aoi$traveltime[[1]])
```

From the syntax above, you can see how you can access a single object
within the nested list column (e.g. by using the list accessor `[[`). In
this case, the shape of the `traveltime` indicator is a single-row and
two-column tibble with the average minutes and the distance category as its value. 
We can now use either of two functions to transform the portfolio to long
or wide formats:

```{r elev-to-wide}
portfolio_long(aoi)
portfolio_wide(aoi)
```

Both function will, by default, automatically detect nested-list columns and 
change the data layout. In this case, the result still has 9 rows just like the 
original data frame because the indicator `traveltime` consisted of just a single
row per asset. 

We could serialize the object to disk in either format by calling `write_portfolio()`
with the respective `format` argument:

```{r simple-export}
dsn_long <- tempfile(fileext = ".gpkg")
dsn_wide <- tempfile(fileext = ".gpkg")
write_portfolio(aoi, dsn_long, format = "long", quiet = TRUE)
write_portfolio(aoi, dsn_wide, format = "wide", quiet = TRUE)
```

## The harder case - indicators with multi-row output

Let's continue to query an indicator which has a multi-row output, i.e.
precipitation statistics from [WorldClim](https://www.worldclim.org/).

```{r calc-prec}
aoi <- get_resources(aoi, get_worldclim_precipitation(years = 2018))
aoi <- calc_indicators(aoi, calc_precipitation_wc(stats = "mean"))
print(aoi)
```

We see that in addition to the `traveltime` indicator, we now obtained an
additional nested-list column called `precipitation_wc`.
Note, however, the differences in the shape of the indicator tibble when we 
take a closer look for a specific asset:

```{r investigate-precip}
print(aoi$precipitation_wc[[1]])
```

For a single asset, we obtain a tibble with 12 rows (for each month in
the queried year 2018). Now, let's have a look at what happens if we transform
this table to long format, this time specifically requesting to only extract
the `precipitation_wc` indicators:

```{r prec-unnest}
portfolio_long(aoi, indicators = "precipitation_wc")
```

Instead of 9 rows, we get a tibble with 108 rows (9 assets \* 12), with
the metadata for each asset with the geometry column and other identifying 
values being repeated 12 times each. For very large portfolios, this data layout 
might be very memory intensive. In these cases it might be more favourable to 
transform the portfolio to a wide layout.

```{r fun-prec-to-wide}
portfolio_wide(aoi, indicators = "precipitation_wc")
```

From the example output above we see that in this case, we obtain a resulting
object with 9 rows only. The indicator data is now found in the respective
columns which are named according to this schema:

`<indicator-name>_<datetime>_<variable>_<unit>`

with the values being found in the rows for each unique combination of this 
pattern.

Note, that the `traveltime` still is represented as a nested-list column. When
serializing to disk, all present indicators are going to extracted in order to
be able to serialize to spatial data formats. If it is not desired to include
certain indicators you will have to subset the portfolio as indicated 
in the following code block:

```{r write-output}
dsn <- tempfile(fileext = ".gpkg")
write_portfolio(select(aoi, traveltime), dsn, quiet = TRUE)
```
