---
title: "Quickstart"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quickstart}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

In the following we will demonstrate an idealized workflow based on a subset
of the Global Forest Watch (GFW) data set that is delivered together with this 
package. You can follow along the code snippets below to reproduce the results. 
Please note that to reduce the time it takes to process this vignette, we will 
not download any resources from the internet. In a real use case, thus processing 
time might substantially increase because resources have to be downloaded and 
real portfolios might be larger than the one created in this example. 

This vignette assumes that you have already followed the steps in 
[Installation](https://mapme-initiative.github.io/mapme.biodiversity/articles/installation.html)
and have familiarized yourself with the terminology used in the package. If you 
are unfamiliar with the terminology used here, please head over to the 
[Terminology](https://mapme-initiative.github.io/mapme.biodiversity/articles/terminology.html)
article to learn about the most important concepts. 

The idealized workflow for using `{mapme.biodiversity}` consists of the following
steps:

- prepare your sf-object containing only geometries of type `'POLYGON'` or `'MULTIPOLYGON'`
- decide which indicator(s) you wish to calculate and make the required resource(s) available
- conduct your indicator calculation, which adds a nested list column to your portfolio object
- continue your analysis in R or decide to export your results to a spatial data format to use it with
  other geospatial software

# Getting started

First, we will load the `{mapme.biodiversity}` and the `{sf}` package for handling
spatial vector data. For tabular data handling, we will also load the `{dplyr}` 
and `{tidyr}` packages. Then, we will read an internal GeoPackage which includes 
part of the geometry of a protected area in the Dominican Republic from the 
WDPA database.

```{r setup, message=FALSE}
library(mapme.biodiversity)
library(sf)
library(dplyr)
library(tidyr)

aoi_path <- system.file("extdata", "gfw_sample.gpkg", package = "mapme.biodiversity")
aoi <- st_read(aoi_path, quiet = TRUE)
aoi
```

```{r simplify-aoi, echo = FALSE}
aoi <- st_simplify(aoi, preserveTopology = TRUE, dTolerance = 500)
```

# Setting standard option

We use the `mapme_options()` function and set some arguments, such as the
output directory, that are important to govern the subsequent processing.
For this, we create a temporary directory. Internally, to save time on
downloading when building this vignette, we copied already existing files
to that output location (code not shown here).

```{r init-data-dir, echo=FALSE}
outdir <- file.path(tempdir(), "mapme-resources")
mapme.biodiversity:::.copy_resource_dir(outdir)
```

```{r portfolio-opts}
outdir <- file.path(tempdir(), "mapme-resources")
dir.create(outdir, showWarnings = FALSE)

mapme_options(
  outdir = outdir,
  verbose = TRUE
)
```

The `outdir` argument points towards a directory on the local file system of your 
machine. All downloaded resources will be written to respective directories nested 
within `outdir`. 

Once you request a specific resource for your portfolio, only those files will be 
downloaded that are missing to match its spatio-temporal extent. This behavior 
is beneficial, e.g. in case you share the `outdir` between different projects to 
ensure that only resources matching your current portfolio are returned.

The `verbose` logical controls whether or not the package
will print informative messages during the calculations. Note, 
that even if set to `FALSE`, the package will inform users about any potential 
errors or warnings.

# Getting the right resources

You can check which indicators are available via the `available_indicators()`
function:

```{r query-indicator}
available_indicators()
available_indicators("treecover_area")
```

Say, we are interested in the `treecover_area` indicator.
We can learn more about this indicator and its required resources by using
either of the commands below or, if you are viewing the online version, head
over to the [treecover_area](https://mapme-initiative.github.io/mapme.biodiversity/reference/treecover_area.html) documentation.

```{r help-indicator, eval = FALSE}
?treecover_area
help(treecover_area)
```

By inspecting the help page we learned that this indicator requires the 
`gfw_treecover` and `gfw_lossyear` resources and it requires to specify three 
extra arguments: the years for which to calculate treecover, the minimum size
of patches to be considered as forest and the minimum canopy coverage of a single
pixel to be considered as forested.

With that information at hand, we can start to retrieve the required resource. 
We can learn about all available resources using the `available_resources()` 
function:

```{r query-resources}
available_resources()
available_resources("gfw_treecover")
```

For the purpose of this vignette, we are going to download both, the `gfw_treecover`
and `gfw_lossyear` resources. We can get more detailed information about a given 
resource, by using either of the commands below to open up the help page. If 
you are viewing the  online version of this documentation, you can simply head 
over to the [gfw_treecover](https://mapme-initiative.github.io/mapme.biodiversity/reference/gfw_treecover.html)
resource documentation.

```{r help-resource, eval = FALSE}
?gfw_treecover
help(gfw_treecover)
?gfw_lossyear
help(gfw_lossyear)
```

We can now make the required resources available for our portfolio. We will
use a common interface that is used for all resources, called `get_resources()`.
We have to specify our portfolio object and supply one or more resource functions
with their respective arguments. This will then download the matching resources
to the output directory specified earlier.

```{r get-gfw}
aoi <- get_resources(
  x = aoi,
  get_gfw_treecover(version = "GFC-2023-v1.11"),
  get_gfw_lossyear(version = "GFC-2023-v1.11")
)
```


# Calculate specific indicators

The next step consists of calculating specific indicators. Note that each 
indicator requires one or more resources that were made available via the 
`get_resources()` function explained above. You will have to re-run this 
function in every new R session, but note that data that is already available
will not be re-downloaded. 

Here, we are going to calculate the `treecover_area` indicator which is based 
on the resources from GFW. Since the resources have been made available in the 
previous step, we can continue requesting the calculation of our desired indicator. 
Note the command below would issue an error in case a required resource has not 
been made available via `get_resources()` beforehand.

```{r calc-indicator}
aoi <- calc_indicators(
  aoi,
  calc_treecover_area(years = 2000:2023, min_size = 1, min_cover = 30)
)
```

Now let's take a look at the results. In addition to the metadata we are 
already familiar with, we see that there is an additional column called
`treecover_area` which contains a `tibble`. 

```{r print-aoi}
aoi
```

The indicator is represented as a nested-list column in our `sf`-object that is 
named alike the requested indicator. For our single asset, this column contains 
a tibble with 6 rows and four columns. Let's have a closer look at this object

```{r print-indicator}
aoi$treecover_area
```

The tibble follows a standard output format, which is the same for all indicators.
Each indicator is represented as a tibble with the four columns `datetime`, 
`variable`, `unit`, and `value`. In case of the `treecover_area` indicator, 
the variable is called `treecover` and is expressed in `ha`.

Let's quickly visualize the results:

```{r plot-treecover, echo = FALSE, warning=FALSE}
mapme_options(verbose = FALSE)
data <- portfolio_long(aoi)

plot(value ~ datetime, data,
  main = "Treecover over time",
  xlab = "Year", ylab = "ha",
  pch = 16, col = "steelblue"
)
```

If you wish to change the layout of an portfolio, you can use `portfolio_long()`
and `portfolio_wide()` (see the respective [online tutorial](https://mapme-initiative.github.io/mapme.biodiversity/articles/output-wide.html)). 
Especially for large portfolios, it is usually a good idea to keep the geometry 
information in a separated variable to keep the size of the data object relatively 
small.

```{r long-drop-geoms}
geoms <- st_geometry(aoi)
portfolio_long(aoi, drop_geoms = TRUE)
```

# Enabling parallel computing

`{mapme.biodiversity}` follows the parallel computing paradigm of the
[`{future}`](https://cran.r-project.org/package=future) package. That means 
that you as a user are in the control if and how you would like to set up 
parallel processing. Since `{mapme.biodiversity} v0.7` supports 
parallel processing on a nested-level. The outer level applies parallel processing 
to the assets in a portfolio, the inner level to potential chunks for polygons 
that are larger than the specified chunks size or the single components
of assets of type `MULTIPOLYGON`. 

The maximum chunk size is specified in hectares via `mapme_options()` and defaults
to 100,000 ha. Polygons larger than this threshold will be split into chunks
of roughly the same size.

Fine-control of parallel processing is given by using future topologies
(find more information [here](https://cran.r-project.org/package=future/vignettes/future-3-topologies.html)).
To process all assets sequentially, but allow to spawn up to 4 workers to 
process chunks in parallel you might specify:

```{r parallel-1, eval = FALSE}
library(future)
plan(list(sequential, tweak(cluster, workers = 6)))
```

As another example, with the code below one would apply parallel processing of 2 
assets, with each having 4 workers available to process chunks, thus requiring 
a total of 8 available cores on the host machine. Be sure to not request
more workers than available on your machine.

```{r parallel, eval = FALSE}
library(progressr)

plan(list(tweak(cluster, workers = 2), tweak(cluster, workers = 4)))

with_progress({
  aoi <- calc_indicators(
    aoi,
    calc_treecover_area(
      min_size = 1,
      min_cover = 30
    )
  )
})

plan(sequential) # close child processes
```

# Exporting an portfolio object

You can use the `write_portfolio()` function to save a processed 
portfolio object to disk as a `GeoPackage`. This allows sharing your data with 
contributors who might not be using R, but any other geospatial software. 
Simply point towards a non-existing file on your local disk to write the portfolio. 
You can use `read_portfolio()` to read back a GeoPackage written in such a way
into R:

```{r write-portfolio}
dsn <- tempfile(fileext = ".gpkg")
write_portfolio(x = aoi, dsn = dsn, quiet = TRUE)
from_disk <- read_portfolio(dsn, quiet = TRUE)
from_disk
```

```{r delete-dsn, echo=FALSE}
file.remove(dsn)
```

