---
title: "Terminology"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Terminology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 5
)
```

Here we present a quick introduction to the most important terminologies and concepts
used throughout this tutorial:

- **Portfolio**: We define a portfolio as
the collection of single assets in an `{sf}` object. The package
does only allow assets with geometries of type `'POLYGON'` to be included in the object.
In case that some of your data has type `'MULTIPOLYGON'` (or others) you are required to 
transform your data via `sf::st_cast()` before initiating your portfolio. 
We also expect the data to be in geographic coordinates. 
In case the CRS of the portfolio is projected, we will attempt to transform to
`EPSG:4326`. Any metadata that is needed for your analysis is allowed
in the columns of the object and will be retained throughout the workflow so that
you can access it once you have calculated your indicators.
The package will add a *nested list columns* for each indicator to the portfolio
`sf`-object. Portfolio-wide arguments that govern the processing workflow, 
are set using the `mapme_options()` config functions.

- **Resources**: A resource is a supported dataset that can be downloaded via the 
`get_resources()` function. A resource represents either a raster or a 
vector data set. Vector datasets are saved to disk as
GeoPackages. Raster datasets can be tiled so that only tiles that cover the 
spatial extent of the portfolio will be downloaded. Global raster resources will
be downloaded completely. Additionally, some raster resources are time-invariant
(e.g. a `nasa_srtm`), others contain a temporal dimension (e.g. `esalandcover`). 
If a raster resource contains a temporal dimension, only those rasters
intersecting with user supplied time-frame will be downloaded. To learn about resources and
their arguments simply query the help-page for that resource (e.g. via `?gfw_treecover`).

- **Indicators**: An indicator calculates a tabular output for all
assets in a portfolio. Every indicator requires one or more resources to be
available locally and will inform you about any missing resources. 
Custom arrguments might be set for some indicators, but sensible defaults 
have been set. The package will inform you about any misspecification of arguments. 
To learn more about a specific indicator and its required arguments, simply query 
the help-page (e.g. via `?treecover_area`). 
When an indicator calculation is done a *nested list column* with the same name 
will be added to the portfolio. For assets where no sensible values can be 
calculated (e.g. for polygons over the ocean and the `treecover_area` indicator),
`NA` values will be inserted.

- **Engines**: Several indicators support different engines for the extraction of
the numeric outputs. Currently, these engines are based on either `terra::zonal()`,
`terra::extract()`, or `exactextractr::exact_extract()`. We took 
great care to set these engines to sensible defaults. However, to 
customize your workflow you can choose to use a different engine instead.

- **Processing Modes**: The processing mode is an important concept if you wish 
to extend the package with a new indicator, but not so much if you simply want to
use it for your statistical analysis. For most indicators the default processing 
mode is equal to `'asset'`, meaning that for each asset in the 
portfolio the required input resources are matched to the spatio-temporal extent 
of that specific asset and then the indicator calculation is conducted. For some 
resources/indicators this is not the most efficient approach. These are most 
commonly very coarse resolution raster resources where a cropping of many 
relatively small assets would lead to an increased overhead and as a result
very long processing times. For these indicators, the default processing mode is
set to `'portfolio'`, meaning that the indicator is first calculated for the
complete spatio-temporal extent of the portfolio and then the extraction is conducted
for every single asset. If you want to open a pull-request for a new indicator for
which neither of these two approaches yields satisfactory performance, please contact
the package maintainers to discuss your use-case in order to evaluate if another
processing mode can be added.
