---
author: "Darius A. Görgen"
title: "Workflow"
date: "Last modified: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 5
)
```

The most important steps of the processing workflow of `{mapme.biodiversity}` 
are described here. A typical workflow consists of the following steps:

-  initialization of a portfolio object from an `{sf}` object together with portfolio-wide parameters
-  identify the required resources needed for the indicators you want to calculate
-  download one or more resources that are required for your desired indicators
-  automated calculation of the indicators for all assets within a portfolio
-  then, you are free to conduct your statistical analysis within R, or,
-  you can export the portfolio as a GeoPackage to share with others or use it with other Geo-spatial software (e.g. QGIS)

```{r wflow, echo = FALSE}
library(DiagrammeR)

mermaid("
  graph TB

    classDef default fill:#c2dddd,stroke-width:2px,stroke:#034858;

    A(Portfolio initialization)
    B(Get resources)
    C(Calculate indicators)
    D(Statistical analysis within R)
    E(Export to GeoPackage)
    A-->B
    B-->C
    C-->D
    C-->E

")
```

