---
author: "Darius A. Görgen"
title: "Workflow"
date: "Last modified: `r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 5
)
```

The most important steps of the processing workflow 
of `{mapme.biodiversity}` are described here. The first crucial step is to initiate a portfolio object with 
all important portfolio-wide parameters (e.g. the temporal extent, input/output structure,
download utility and parallel processing). In a next step the required resources
for the analysis are downloaded. Once all required resources are made available, 
supported indicators are calculated for each asset within the portfolio. Then,
users can either continue their data analysis within R or decide to export
the portfolio to a GeoPackage and continue to work with other Geo-spatial software.
The relevant steps (see figure) in a `{mapme.biodiversity}` workflow thus comprise:

-  initialization of a portfolio object from an `{sf}` object together with portfolio-wide parameters
-  download of a selection of resources via a standardized interface
-  calculation of indicators for all assets within a portfolio via a standardized interface
-  then, users are free to conduct their statistical analysis within R
-  optionally, export the portfolio as a GeoPackage to share with others or use other Geo-spatial software

```{r wflow, echo = FALSE}
library(DiagrammeR)

mermaid("
  graph TB

    classDef default fill:#c2dddd,stroke-width:2px,stroke:#034858;

    A(Portfolio initialization)
    B(Get resources)
    C(Calculate indicators)
    D(Statistical analysis within R)
    E(Export to GeoPackage)
    A-->B
    B-->C
    C-->D
    C-->E

")
```

